### 2022-11-11 submission/manuscript.Rmd
First full version of Discussion

Over the past decade, the question of whether to rarefy microbial community sequence data has been controversial. The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sampling effort when comparing communities based on alpha and beta diversity indices. Compared to all other methods, rarefaction removed the correlation between sequencing depth and alpha or beta diversity metrics when the sequencing depth varied by as much as 97-fold across samples. I showed that this correlation could lead falsely detecting differences between treatment groups if sampling depth and sequencing effort are confounded [e.g., @Morris2013]. The correlation with sampling effort leads to an artificial increase in the variation between samples and a reduced power to detect true differences in alpha and beta diversity. For these reasons, rarefaction is a valuable tool to control for uneven sampling effort until improved statistical procedures are developed or it becomes possible to more evenly distribute sequencing effort across samples.

@@

Two common critique of rarefaction is that the approach ommits valid data and that the selection of the value to rarefy the sequence data to is arbitrary [@McMurdie2014]. I disagree with both critiques. All of the data are used to calculate the mean value of the rarefied metrics; none of it is excluded. When the dataset is subsampled, every sequence has a random chance of being included in the calculated metric. When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risk of oversampling rare taxa is mititgated. As for the second critism, I would resist the claim that the selection of the rarefaction threshold is arbitrary. In practice, there is a tradeoff between sampling breadth and depth. Greater breadth will increase the statistical power to compare treatment groups and greater depth will increase the resolution to describe the communities. My process for picking a break involves looking for a natural break in the distribution of the number of sequences. For example, the Lake dataset used in this study had a clear break around 10,000 sequences. I would also consider what samples are below any break that I select. If there were critical samples below the break I would either reduce the break point or I would generate more sequences from those samples. Furthermore, regardless of the break point one selects, they could repeat an analysis at multiple thresholds to assess the tradeoff between statistical power and resolution. As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sampling depth by 10 fold to increase the coverage by several percentage points. I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistica power over depth and resolution.

* Need better methods of pooling libraries that result in more even distribution of sequences across samples
* We don't have true absolute numbers
* Rarefy your data

@@

**Null treatment model.** I created an unbiased and biased treatment model. In the unbiased model, samples were randomly assigned to one of two treatment groups. In the biased treatment model, samples that had more than the median number of seqeunces for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group. Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\upalpha$) of 0.05.

>>>>

Over the past decade, the question of whether to rarefy microbial community sequence data has become controversial. The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sampling effort when comparing communities based on alpha and beta diversity indices. Compared to all other methods, rarefaction removed the correlation between sequencing depth and alpha or beta diversity metrics when the sequencing depth varied by as much as 97-fold across samples. I showed that this correlation could lead falsely detecting differences between treatment groups if sampling depth and sequencing effort are confounded [e.g., @Morris2013]. The correlation with sampling effort leads to an artificial increase in the variation between samples and a reduced power to detect true differences in alpha and beta diversity. For these reasons, rarefaction is a valuable tool to control for uneven sampling effort until improved statistical procedures are developed or it becomes possible to more evenly distribute sequencing effort across samples.

@@

Two common critique of rarefaction is that the approach ommits valid data and that the selection of the value to rarefy the sequence data to is arbitrary [@McMurdie2014]. I disagree with both critiques. All of the data are used to calculate the mean value of the rarefied metrics; none of it is excluded. When the dataset is subsampled, every sequence has a random chance of being included in the calculated metric. When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risks of ignoring or oversampling rare taxa are mititgated. Furthermore, it is curious that the study making the original critique removed sequences that did not appear in at least three samples or had an abundance of one in any sample. A parallel analysis to this study has demonstrated that many of these sequences are likely valid and that removal of rare sequences can bias alpha and beta diversity metrics and reduce statistical power [@Schloss2020]. As for the second critism, I would resist the claim that the selection of the rarefaction threshold is arbitrary. In practice, there is a tradeoff between sampling breadth and depth. Greater breadth will increase the statistical power to compare treatment groups and greater depth will increase the resolution to describe the communities. My process for picking a break involves looking for a natural break in the distribution of the number of sequences. For example, the Lake dataset used in this study had a clear break around 10,000 sequences. I would also consider what samples are below any break that I select. If there were critical samples below the break I would either reduce the break point or I would generate more sequences from those samples. Furthermore, regardless of the break point one selects, they could repeat an analysis at multiple thresholds to assess the tradeoff between statistical power and resolution. As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sampling depth by 10 fold to increase the coverage by several percentage points. I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistical power over depth and resolution.

The up to 100-fold difference in sample sizes is an unfortunate byproduct of how seqencing libraries are constructed. Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequencs that all them to later identify which sample each sequence belongs. When the PCRs are pooled efforts are taken to pool the fragments in equimolar ratios. Researchers use one of two approaches. First, they often will quantify the concentration of DNA from each PCR and then pool DNA in the desired amounts. Alternatively, they may use normalization plates where each well can hybridize a uniform amount of DNA that is then eluted and pooled. Clearly, both approaches have limitations that reduce the ability to truly achieve equimolar mixture. In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons [@Morris2013]. Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.

All simulations have weaknesses and should be interpreted with caution. However, the simulated communites generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups. For now, is is perfectly admissible and proper to rarefy microbial community data or risk reaching unwarrented conclusions.

@@

**Null treatment models.** I created an unbiased and biased treatment model. In the unbiased model, samples were randomly assigned to one of two treatment groups. In the biased treatment model, samples that had more than the median number of seqeunces for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group. Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\upalpha$) of 0.05.

### 2022-11-11 submission/manuscript.tex
First full version of Discussion

Over the past decade, the question of whether to rarefy microbial
community sequence data has been controversial. The analyses I presented
here strongly indicate that rarefaction is necessary to control for
uneven sampling effort when comparing communities based on alpha and
beta diversity indices. Compared to all other methods, rarefaction
removed the correlation between sequencing depth and alpha or beta
diversity metrics when the sequencing depth varied by as much as 97-fold
across samples. I showed that this correlation could lead falsely
detecting differences between treatment groups if sampling depth and
sequencing effort are confounded (e.g., 1). The correlation with
sampling effort leads to an artificial increase in the variation between
samples and a reduced power to detect true differences in alpha and beta
diversity. For these reasons, rarefaction is a valuable tool to control
for uneven sampling effort until improved statistical procedures are
developed or it becomes possible to more evenly distribute sequencing
effort across samples.

@@

Two common critique of rarefaction is that the approach ommits valid
data and that the selection of the value to rarefy the sequence data to
is arbitrary (5). I disagree with both critiques. All of the data are
used to calculate the mean value of the rarefied metrics; none of it is
excluded. When the dataset is subsampled, every sequence has a random
chance of being included in the calculated metric. When that subsampling
is repeated a large number of times (e.g., 100 or 1000) the risk of
oversampling rare taxa is mititgated. As for the second critism, I would
resist the claim that the selection of the rarefaction threshold is
arbitrary. In practice, there is a tradeoff between sampling breadth and
depth. Greater breadth will increase the statistical power to compare
treatment groups and greater depth will increase the resolution to
describe the communities. My process for picking a break involves
looking for a natural break in the distribution of the number of
sequences. For example, the Lake dataset used in this study had a clear
break around 10,000 sequences. I would also consider what samples are
below any break that I select. If there were critical samples below the
break I would either reduce the break point or I would generate more
sequences from those samples. Furthermore, regardless of the break point
one selects, they could repeat an analysis at multiple thresholds to
assess the tradeoff between statistical power and resolution. As shown
in my analysis of Good's coverage values, most studies obtain an ample
level of coverage and would need to increase their sampling depth by 10
fold to increase the coverage by several percentage points. I contend
that for most studies increased sequencing effort should be applied to
improve sampling breadth and statistica power over depth and resolution.

\begin{itemize}
\tightlist
\item
  Need better methods of pooling libraries that result in more even
  distribution of sequences across samples
\item
  We don't have true absolute numbers
\item
  Rarefy your data
\end{itemize}

@@

\textbf{Null treatment model.} I created an unbiased and biased
treatment model. In the unbiased model, samples were randomly assigned
to one of two treatment groups. In the biased treatment model, samples
that had more than the median number of seqeunces for a dataset were
assigned to one treatment group and the rest were assigned to a second
treatment group. Comparison of any diversity metric across the two
treatment groups should have only yielded a significant result in 5\% of
the simulations when testing under a Type I error (i.e., \(\upalpha\))
of 0.05.

>>>>

Over the past decade, the question of whether to rarefy microbial
community sequence data has become controversial. The analyses I
presented here strongly indicate that rarefaction is necessary to
control for uneven sampling effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction removed the correlation between sequencing depth and alpha
or beta diversity metrics when the sequencing depth varied by as much as
97-fold across samples. I showed that this correlation could lead
falsely detecting differences between treatment groups if sampling depth
and sequencing effort are confounded (e.g., 1). The correlation with
sampling effort leads to an artificial increase in the variation between
samples and a reduced power to detect true differences in alpha and beta
diversity. For these reasons, rarefaction is a valuable tool to control
for uneven sampling effort until improved statistical procedures are
developed or it becomes possible to more evenly distribute sequencing
effort across samples.

@@

Two common critique of rarefaction is that the approach ommits valid
data and that the selection of the value to rarefy the sequence data to
is arbitrary (5). I disagree with both critiques. All of the data are
used to calculate the mean value of the rarefied metrics; none of it is
excluded. When the dataset is subsampled, every sequence has a random
chance of being included in the calculated metric. When that subsampling
is repeated a large number of times (e.g., 100 or 1000) the risks of
ignoring or oversampling rare taxa are mititgated. Furthermore, it is
curious that the study making the original critique removed sequences
that did not appear in at least three samples or had an abundance of one
in any sample. A parallel analysis to this study has demonstrated that
many of these sequences are likely valid and that removal of rare
sequences can bias alpha and beta diversity metrics and reduce
statistical power (3). As for the second critism, I would resist the
claim that the selection of the rarefaction threshold is arbitrary. In
practice, there is a tradeoff between sampling breadth and depth.
Greater breadth will increase the statistical power to compare treatment
groups and greater depth will increase the resolution to describe the
communities. My process for picking a break involves looking for a
natural break in the distribution of the number of sequences. For
example, the Lake dataset used in this study had a clear break around
10,000 sequences. I would also consider what samples are below any break
that I select. If there were critical samples below the break I would
either reduce the break point or I would generate more sequences from
those samples. Furthermore, regardless of the break point one selects,
they could repeat an analysis at multiple thresholds to assess the
tradeoff between statistical power and resolution. As shown in my
analysis of Good's coverage values, most studies obtain an ample level
of coverage and would need to increase their sampling depth by 10 fold
to increase the coverage by several percentage points. I contend that
for most studies increased sequencing effort should be applied to
improve sampling breadth and statistical power over depth and
resolution.

The up to 100-fold difference in sample sizes is an unfortunate
byproduct of how seqencing libraries are constructed. Researchers
perform separate PCRs for each sample with unique index (aka, barcode)
sequencs that all them to later identify which sample each sequence
belongs. When the PCRs are pooled efforts are taken to pool the
fragments in equimolar ratios. Researchers use one of two approaches.
First, they often will quantify the concentration of DNA from each PCR
and then pool DNA in the desired amounts. Alternatively, they may use
normalization plates where each well can hybridize a uniform amount of
DNA that is then eluted and pooled. Clearly, both approaches have
limitations that reduce the ability to truly achieve equimolar mixture.
In addition, for some samples it is common to co-amplify non-specific
DNAs which may add to the challenges of achieving equimolar mixtures of
the desired amplicons (1). Regardless, it is clear that better
strategies are needed to reduce the variation in the number of sequences
generated for each sample.

All simulations have weaknesses and should be interpreted with caution.
However, the simulated communites generated and analyzed in this study
had the advantage of being designed with known properties including the
alpha and beta diversity and the their differences between treatment
groups. For now, is is perfectly admissible and proper to rarefy
microbial community data or risk reaching unwarrented conclusions.

@@

\textbf{Null treatment models.} I created an unbiased and biased
treatment model. In the unbiased model, samples were randomly assigned
to one of two treatment groups. In the biased treatment model, samples
that had more than the median number of seqeunces for a dataset were
assigned to one treatment group and the rest were assigned to a second
treatment group. Comparison of any diversity metric across the two
treatment groups should have only yielded a significant result in 5\% of
the simulations when testing under a Type I error (i.e., \(\upalpha\))
of 0.05.
